# Benchmark

# imports
```{r message=FALSE}
library(Seurat)
library(scregclust)
library(glue)
```
#

# functions
```{r  message=FALSE}

sample_data <- function(data, fraction, replace){
  # genes x cells
  dataset <- data
  N <- round(ncol(dataset)*fraction)
  
  column_indices <- sample(ncol(dataset), size = N, replace = replace)
  
  sample <-  dataset[,column_indices]
  
  sample
}


process_data <- function(pbmcData){
  
  pbmc <- CreateSeuratObject(
  counts = pbmcData, min.cells = 3, min.features = 200
  )

  pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT.")
  pbmc <- subset(pbmc, subset = percent.mt < 30 & nFeature_RNA < 6000)
  pbmc <- SCTransform(pbmc, variable.features.n = 6000)

  z <- GetAssayData(pbmc, layer = 'scale.data')
  
  z
  
}

initialize <- function(assay, mode){

  out <- scregclust_format(assay, mode)
  out
}

# Helpers

find_gene_cluster_and_regulators <- function(gene_symbol, fit){

  results <- fit$results[[1]]
  print('--- Taking results for first penalty ---')
  
  is_regulator <- results$is_regulator
  
  gene_symbols <- results$genesymbols
  
  non_regulators_gene_symbols <- gene_symbols[is_regulator==0]
  
  index_absolute <- which(gene_symbols == gene_symbol)
  
  outputs <- results$output
  
  modes <- length(outputs)
  
  print(glue('--- Algo found {modes} modes and started cycling ---'))
  
  answers <- list()
  
  for (i in seq_len(modes)){
  
    sub_output <- outputs[[i]]
    
    i_cluster <- sub_output$cluster[[index_absolute]]
    
    i_cluster_all <- sub_output$cluster_all[[index_absolute]]
    
    answers[[as.character(i)]] = list(
    
    cluster = i_cluster,
    
    cluster_all = i_cluster_all)
    
    regulators <- list()
    
    if (!is.na(i_cluster_all)){
    
      clus <- max(i_cluster_all, i_cluster)
      
      
      index_non_regulator <- which(non_regulators_gene_symbols == gene_symbol)
      
      
      clus_regulators_gene_symbol <- gene_symbols[is_regulator == 1][sub_output$models[, clus]]
      
      number_regulators <- length(clus_regulators_gene_symbol)
      
      
      for (j in seq_len(number_regulators)){
      
        gene_coeff <- sub_output$coeffs[[clus]][j, index_non_regulator]
        
        if (gene_coeff != 0L){
          
          regulators <- append(regulators, clus_regulators_gene_symbol[[j]])
        }
      }
      
      answers[[as.character(i)]][["regulators"]] = regulators
    
    }
    
    else {
    
      answers[[as.character(i)]][["regulators"]] = NA
    }
  }
  
  answers
  
  }


```

#

# data
```{r  message=FALSE}
pbmcData <- Read10X_h5(
  'pbmc_granulocyte_sorted_3k_filtered_feature_bc_matrix.h5',
  use.names = TRUE,
  unique.features = TRUE
)[["Gene Expression"]]

```
#

# Get benchmark fits
```{r}
versions <- c(1, 2, 3, 4, 5)

N_rounds <- 20

sampling_frac <- 0.5

with_replacement <- FALSE

penalties <- c(0.1, 0.2, 0.3, 0.4, 0.5)

set.seed(8374)

for (p in seq_len(length(penalties))){
  for (i in seq_len(N_rounds)){
    print(glue("\n--- round {i} starting ---\n"))
    
    sample <- sample_data(pbmcData, fraction = sampling_frac, replace = with_replacement)
    
    assay <- process_data(sample)
    print(glue("\n--- {dim(assay)[2]} cells used ---\n"))
    
    out <- initialize(assay, 'TF') 
    
  
    fit <- scregclust(assay, 
                      out$genesymbols, 
                      out$is_regulator, 
                      penalization = penalties[p],
                      n_cl = 10L, 
                      n_cycles = 50L, 
                      noise_threshold = 0.05, 
                      center=FALSE,
                      sample_assignment = out$sample_assignment
     )
    
    version <- versions[p]
    saveRDS(fit, file = glue("test_run/v{version}_sample_{i}_{penalties[p]}_fit.rds"))
    
    rm(fit)
  }
 
  }
```











#

# Format results
```{r}
folder_path <- "benchmark_models"

files <- list.files(folder_path)

df <- data.frame()

penalties <-c(0.1, 0.2, 0.3, 0.4, 0.5)

for (file in files){
  
  i_fit <- readRDS(glue('{folder_path}/{file}'))
  
  
  gene_symbols <- i_fit$results[[1]]$genesymbols
  
  
  modes <- length(i_fit$results[[1]]$output)
  
  
  for (i in seq_len(modes)){
  
    for (gene in gene_symbols){
    
      results <- find_gene_cluster_and_regulators(gene, i_fit)[[i]]
      
      regulators <- as.character(results[["regulators"]])
      
      if (is.na(results[["cluster"]])){
        
        bool_is_regulator <- TRUE
        
      }
        
      else {
        
       bool_is_regulator <- FALSE
       
      }
    
      times <- length(regulators)
      
      if (times==0){
      
        times <- 1
        
        regulators <- c(NA)
        
      }
      
      df_temp <- data.frame(
                            iteration = c(rep(strsplit(file, "_")[[1]][3], times = times)),
                            penalty = c(rep(penalties[as.integer(substring(file, 2, 2))], times = times)),
                            state = c(rep(i, times = times)),
                            gene = c(rep(gene, times = times)),
                            is_regulator = c(rep(bool_is_regulator, times = times)),
                            cluster = c(rep(results[["cluster"]], times = times)),
                            cluster_all = c(rep(results[["cluster_all"]], times = times)),
                            regulators = c(regulators)
                            )
      
      df <- rbind(df, df_temp)
  }
  
    }
  
  export_name <- strsplit(file,"rds")[[1]]
  write.csv(df, glue("benchmark_models_csv/{export_name}csv"), row.names=FALSE)
  
  rm(df)
  
}


```




















# 

# data
```{r  message=FALSE}
# Load data - wget is not working
pbmcData <- Read10X_h5(
  'pbmc_granulocyte_sorted_3k_filtered_feature_bc_matrix.h5',
  use.names = TRUE,
  unique.features = TRUE
)[["Gene Expression"]]

```
#



# benchmark
```{r}
version <- 0

N_rounds <- 20

sampling_frac <- 0.5

with_replacement <- FALSE

penalization <- 0.2

set.seed(8374)

for (i in seq_len(N_rounds)){
  print(glue("\n--- round {i} starting ---\n"))
  
  sample <- sample_data(pbmcData, fraction = sampling_frac, replace = with_replacement)
  
  assay <- process_data(sample)
  print(glue("\n--- {dim(assay)[2]} cells used ---\n"))
  
  out <- initialize(assay, 'TF') 
  

  fit <- scregclust(assay, 
                    out$genesymbols, 
                    out$is_regulator, 
                    penalization = penalization,
                    n_cl = 10L, 
                    n_cycles = 50L, 
                    noise_threshold = 0.05, 
                    center=FALSE,
                    sample_assignment = out$sample_assignment
   )
  
  
  saveRDS(fit, file = glue("tested_models/v{version}_sample_{i}_fit.rds"))
  
  rm(fit)
 
  }
```



```{r}



# Read the .rds file
test <- readRDS('tested_models/v0_sample_2_fit.rds')

write_csv(data, "test.csv")


```











































---
title: "Demonstration of workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pbmc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The methods below are described in our article

> Larsson I & Held F, et al. (2023) Reconstructing the regulatory programs
> underlying the > phenotypic plasticity of neural cancers. Preprint available
> at [bioRxiv](https://www.biorxiv.org/content/10.1101/2023.03.10.532041v1);
> 2023.03.10.532041.

Here we demonstrate the scregclust workflow using the PBMC data from
10X Genomics (available [here](https://www.10xgenomics.com/resources/datasets/pbmc-from-a-healthy-donor-granulocytes-removed-through-cell-sorting-3-k-1-standard-2-0-0).
This is the same data used in an [introductory vignette](https://satijalab.org/seurat/articles/pbmc3k_tutorial) for the
Seurat package. We use [Seurat](https://satijalab.org/seurat/) for
pre-processing of the data.

```{r load-packages, results='hide', message=FALSE}
# Load required packages
library(Seurat)
library(scregclust)
```

































#





# Download the data

We are focusing here on the filtered feature barcode matrix available as an
HDF5 file from the website linked above. The data can be downloaded manually
or using R.

However you obtain the data, the code below assumes that the HDF5 file
containing it is placed in the same folder as this script with the name
`pbmc_granulocyte_sorted_3k_filtered_feature_bc_matrix.h5`.

```{r download-data}
url <- paste0(
  "https://cf.10xgenomics.com/samples/cell-arc/2.0.0/",
  "pbmc_granulocyte_sorted_3k/",
  "pbmc_granulocyte_sorted_3k_filtered_feature_bc_matrix.h5"
)
path <- "pbmc_granulocyte_sorted_3k_filtered_feature_bc_matrix.h5"

download.file(url, path, cacheOK = FALSE, mode = "wb")
```

# Load the data in Seurat and preprocess

To perform preprocessing use Seurat to load the data. The file ships with
two modalities, "Gene Expression" and "Peaks". We only use the former.

```{r load-h5}
pbmcData <- Read10X_h5(
  'pbmc_granulocyte_sorted_3k_filtered_feature_bc_matrix.h5',
  use.names = TRUE,
  unique.features = TRUE
)[["Gene Expression"]]
```

We create a Seurat object and follow the Seurat vignette to subset the
cells and features (genes).

```{r create-seurat-object}
pbmc <- CreateSeuratObject(
  counts = pbmcData, min.cells = 3, min.features = 200
)

pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT.")
pbmc <- subset(pbmc, subset = percent.mt < 30 & nFeature_RNA < 6000)
```

[SCTransform](https://satijalab.org/seurat/articles/sctransform_vignette) is
used for variance stabilization of the data and Pearson residuals for the
6000 most variable genes are extracted as matrix `z`.

```{r apply-var-stabilization, cache=TRUE, message=FALSE}
pbmc <- SCTransform(pbmc, variable.features.n = 6000)

z <- GetAssayData(pbmc, layer = 'scale.data')
dim(z)
```

# Use scregclust for clustering

We then use `scregclust_format` which extracts gene symbols from the
expression matrix and determines which genes are considered regulators.
By default, transcription factors are used as regulators. Setting `mode`
to `"kinase"` uses kinases instead of transcription factors. A list of the
regulators used internally is returned by `get_regulator_list()`.

```{r prep-scregclust}
out <- scregclust_format(z, mode = "TF")
```

`sample_assignment` is initialized to be a vector of `1`s of length `ncol(z)`
and can be filled with a known sample grouping. Here, we do not use it and
just keep it uniform across all cells.

```{r extract-scregclust-arguments}
genesymbols <- out$genesymbols
sample_assignment <- out$sample_assignment
is_regulator <- out$is_regulator
```

Run `scregclust` with number of initial clusters set to 10 and test
several penalties. `noise_threshold` controls the minimum $R^2$ a gene has
to achieve across clusters. Otherwise it is placed in the noise/rag-bag cluster.
The run can be reproduced with the command below. A fitted model can be
downloaded from [GitHub](https://github.com/sven-nelander/scregclust/raw/main/datasets/pbmc_scregclust.rds)
for convenience.

```{r run-scregclust}
set.seed(8374)
fit <- scregclust(
   z, genesymbols, is_regulator, penalization = seq(0.1, 0.5, 0.1),
   n_cl = 10L, n_cycles = 50L, noise_threshold = 0.05, center=FALSE,
   sample_assignment = sample_assignment
 )

#url <- paste0(
#  "https://github.com/sven-nelander/scregclust/raw/main/datasets/",
#  "pbmc_scregclust.rds"
#)
#path <- "pbmc_scregclust.rds"
#download.file(url, path)
#fit <- readRDS("pbmc_scregclust.rds")
```

# Analysis of results

Results can be visualized easily using built-in functions.
Metrics for helping in choosing an optimal penalty can be plotted by calling
`plot` on the object returned from `scregclust`.

```{r viz-metrics, fig.width=7, fig.height=4, fig.dpi=100}
plot(fit)
```

The results for each penalization parameter are placed in a list, `results`,
attached to the `fit` object. So `fit$results[[1]]` contains the results
of running `scregclust` with `penalization = 0.1`. For each penalization
parameter, the algorithm might end up finding multiple optimal clusterings.
The results for each such clustering are contained in the list `output`.
This means that `fit$results[[1]]$output[[1]]` contains the results for
the first final clustering. More than one may be available.

```{r n-configs}
sapply(fit$results, function(r) length(r$output))
```

In this example, two final configurations were found for the lowest
penalization and only a single configuration for all other penalization
parameters.

To plot the regulator network of the first clustering for `penalization = 0.1`
the function `plot_regulator_network` can be used.

```{r viz-reg-network, fig.width=7, fig.height=7, fig.dpi=100}
plot_regulator_network(fit$results[[1]]$output[[1]])
```